---
title: "Encrypting a single file"
author: "John C. Nash"
date: '`r Sys.Date()`'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Abstract

This document gives some suggestions on how to encrypt, decrypt and view a single
(text) file in different computing environments.

# Introduction

Encryption is a key aspect of modern computing and communications. In this article,
we will focus on the encryption of a single **file** of data, normally assumed
to be ordinary text, as well as its decryption and viewing. We will usually 
restrict our attention to **symmetric** encryption, where the same **password**
or **passphrase** is used to encrypt and decrypt the file. However, there are
a number of important tools that use **passkeys** and may use asymmetric or
public-private key pairs.

Note that encryption can be applied to whole storage devices, volumes, directories
or streams. Here we will not explore those possibilities, even if our tools could
be used for such applications.

# Required tools

To simplify our discussion, we will give our tools the *generic* names

- **ecrypt**, a program that encrypts a file from FILE to FILE.CCC
- **dcrypt**, a program that decrypts a file from FILE.CCC to FILE
- **vcrypt**, a program that displays FILE.CCC or part thereof for human reading,
  hopefully removing all readable traces of the decrypted information as it exits.
  This facility is not available for all encryption tools, so the user may have
  to take measures to avoid leaving unencrypted information accessible to those
  not authorized to view it.
  
# Some possible tools

## ccrypt

**ccrypt** (\url{http://ccrypt.sourceforge.net/}) is intended for file encryption.
I happen to be acquainted with its author, Peter Seligman of Dalhousie University.
It is available for most platforms. 

 - ecrypt becomes
 
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; ccrypt FILE
 
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; and outputs FILE.cpt
 
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; The password must be entered twice, though there are options to include\
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; it on the command line, but such options are discouraged for reasons of\
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; security.
 
 - dcrypt becomes (the user is prompted for the password)
         
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; ccrypt -d FILE.cpt
      
 - vcrypt becomes
 
 &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; ccat FILE.cpt

All the above are executed in a terminal (or command) window on Linux, Windows
and (we presume) Macintosh. Note the **qccrypt** GUI wrapper for Linux and
Windows.

### Linux


**ecrypt-ccrypt.sh**


```{r file="enctools/ecrypt-ccrypt.sh", echo=TRUE, eval=FALSE}
```

**dcrypt-ccrypt.sh**


```{r file="enctools/dcrypt-ccrypt.sh", echo=TRUE, eval=FALSE}
```

**vcrypt-ccrypt.sh**


```{r file="enctools/vcrypt-ccrypt.sh", echo=TRUE, eval=FALSE}
```

There are variants of the GUI wrapper **qccrypt** for several Linux distros as
well as source code at http://qccrypt.free.fr/download.html .

### Windows

Note that these tools are intended to be run inside a command terminal screen.
However, there are 32-bit installable and portable Windows packages at, respectively,
http://qccrypt.free.fr/download/qccrypt-0.9.1.msi and 
http://qccrypt.free.fr/download/qccrypt-0.9.1-win32.zip .

**ecrypt-ccrypt.bat**

```{r file="enctools/ecrypt-ccrypt.bat", echo=TRUE, eval=FALSE}
```

**dcrypt-ccrypt.bat**


```{r file="enctools/dcrypt-ccrypt.bat", echo=TRUE, eval=FALSE}
```

**vcrypt-ccrypt.bat**


```{r file="enctools/vcrypt-ccrypt.bat", echo=TRUE, eval=FALSE}
```

### Android

There is a tarball of an Android program 

https://ccrypt.sourceforge.net/download/1.10/ccrypt-1.10.android.tar.gz

but I have NOT determined how to install and use this.

## 7zip

**7zip** (\url{https://7-zip.org/}) is free and open source software intended 
primarily for compressed archiving of collections of files. However, it includes
strong AES-256 encryption in 7z and ZIP formats. Moreover, it is available for
almost all platforms, including Android. However, vcrypt seems unavailable in a 
form that is "view only".

### Linux

**ecrypt-7z.sh**


```{r file="enctools/ecrypt-7z.sh", echo=TRUE, eval=FALSE}
```

**dcrypt-7z.sh**


```{r file="enctools/dcrypt-7z.sh", echo=TRUE, eval=FALSE}
```

### Windows


**ecrypt-7z.bat**


```{r file="enctools/ecrypt-7z.bat", echo=TRUE, eval=FALSE}
```

**dcrypt-7z.bat**

```{r file="enctools/dcrypt-7z.bat", echo=TRUE, eval=FALSE}
```

### Mac ??

### Android

I have found two "apps" on Android that I have verified can unpack password 
protected zip files using the encryption of the 7z family of software:
**7zipper** and **Zarchive**. I will NOT provide 
programs to encrypt and decrypt for Android, as "Android" is actually several
different operating systems in my opinion, and I find the various "help" postings
on the Internet and in Youtube never fit my own devices (two tablets and two
phones, all slightly different). A particular annoyance is that Zarchiver does
not allow decryption of files on a plug-in SD memory card for some reason. There
are a number of postings and videos on how to "fix" this, but I did not find
any worked. If I copied a zip archive to the Internal memory device of my tablet,
then Zarchiver would decrypt it so a text editor (I used the Jota Text Editor)
could view the file.

Two particular concerns with mobile phones and tablets is that the decrypted
(i.e., plaintext) document is written to the storage of the device. This means

- for security, it must be properly erased. This means over-writing the 
storage after the control entry has been modified, as the actual data will 
still be present otherwise;

- with the almost certain network connectivity, it is difficult to be sure that
external agents cannot view the plaintext material before we can erase it. I
do not have the expertise to estimate this risk, but it is certainly not
zero.



## gpg

 - the command for ecrypt becomes
 
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; gpg --symmetric FILE
 
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; and outputs FILE.gpg
 
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; The password must be entered twice, though there are options to include\
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; it on the command line, but such options are discouraged for reasons of\
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; security.
 
 - the command for dcrypt becomes (the user is prompted for the password)
         
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; gpg --decrypt FILE.gpg
      
 - There does not appear to be a reasonable option with gpg for vcrypt 
 


### Linux


**ecrypt-gpg.sh**


```{r file="enctools/ecrypt-gpg.sh", echo=TRUE, eval=FALSE}
```

**dcrypt-gpg.sh**

```{r file="enctools/dcrypt-gpg.sh", echo=TRUE, eval=FALSE}
```

### Windows

Note that these tools are intended to be run inside a command terminal screen.

**ecrypt-gpg.bat**

```{r file="enctools/ecrypt-gpg.bat", echo=TRUE, eval=FALSE}
```

**dcrypt-gpg.bat**


```{r file="enctools/dcrypt-gpg.bat", echo=TRUE, eval=FALSE}
```

## Strategies for use

The fact that different platforms have generally different modes of use forces 
consideration of which tools to use for file encryption in a particular situation.
Some of the questions a user must answer are as follows:

- Will files encrypted by a particular tool require decryption on a different
platform. This is particularly important, for example, in the case of ccrypt,
since it is not reasonably available on Android (and we do not know about
iPhone)